---
title: "Tutorial"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

This tutorial will guide you through the main functions of the `RightSeuratTools` package. The package is designed to provide complementary tools to the `Seurat` package, such as plotting, data manipulation, and more. The package is still under development, and new functions will be added in the future.

## Installation

You can install the `RightSeuratTools` package from GitHub using the `devtools` package. If you don't have the `devtools` package installed, you can install it using the following command:

`install.packages("devtools")`

Once you have `devtools` installed, you can install the `RightSeuratTools` package, as well as another package called `SeuratData` which will be used to load the dataset, using the following command:

`devtools::install_github("Alexis-Varin/RightSeuratTools")`

`devtools::install_github("satijalab/seurat-data")`

## Loading the package

```{r}
suppressPackageStartupMessages(library(RightSeuratTools))
```

## Data loading and preprocessing

For this tutorial, we will use the `pbmc3k` dataset, which contains single-cell RNA-seq data from peripheral blood mononuclear cells (PBMCs).

We will also build on `Seurat`'s tutorial for `pbmc3k` to preprocess the data, available for reference [here](https://satijalab.org/seurat/articles/pbmc3k_tutorial).

```{r}
suppressPackageStartupMessages(library(Seurat))
suppressPackageStartupMessages(library(SeuratData))

# Load the pbmc3k dataset
suppressWarnings(InstallData("pbmc3k"))
data(pbmc3k)
# To be able to use it, we first need to update it to the latest Seurat version
pbmc3k <- UpdateSeuratObject(pbmc3k)
pbmc3k

VlnPlot(pbmc3k, features = c("nFeature_RNA", "nCount_RNA"), ncol = 2)
```

Since the object seems to have already been preprocessed to some extent, as it contains a 'data' layer and we are unsure of which other functions have been used, we will start by testing our first function from `RightSeuratTools` to start from scratch, Right_DietSeurat(), which will remove all the layers and slots from the object, leaving only the counts layer from the RNA assay. We will also choose to remove all meta.data columns except for the 'orig.ident' column.

```{r}
pbmc <- Right_DietSeurat(pbmc3k)
pbmc
```
