---
title: "Getting Started with RightSeuratTools"
output: html_document
vignette: >
  %\VignetteIndexEntry{Getting Started with RightSeuratTools}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

This tutorial will guide you through the main functions of the `RightSeuratTools` package, designed to provide complementary tools to the `Seurat` package, such as plotting, data manipulation, and more. The package is still under development, and new functions will be added in the future.

## Installation

You may install the `RightSeuratTools` package from GitHub using the `devtools` package, which, if you don't have it, may be installed using the following command:

```{r devtools, eval = FALSE}
install.packages("devtools")
```

Once installed, you may install the `RightSeuratTools` package, as well as another package called `SeuratData` which will be used to load the dataset, using the following command:

```{r install, eval = FALSE}
devtools::install_github("Alexis-Varin/RightSeuratTools")
```

```{r seuratdata, eval = FALSE}
devtools::install_github("satijalab/seurat-data")
```

## Data loading

For this tutorial, we will use the `pbmc3k` dataset, which contains single-cell RNA-seq data from peripheral blood mononuclear cells (PBMCs), and made available by [10X Genomics](https://www.10xgenomics.com/datasets/3-k-pbm-cs-from-a-healthy-donor-1-standard-1-1-0).

We will also build on `Seurat`'s tutorial for the `pbmc3k` dataset to preprocess the data, available for reference [here](https://satijalab.org/seurat/articles/pbmc3k_tutorial).

Let's start by loading all the packages we will be using in this vignette:

```{r lib, warning = FALSE, message = FALSE}
library(RightSeuratTools)
library(Seurat)
#library(SeuratData)
library(ggplot2)
```

Next, we will load the `pbmc3k` dataset from `SeuratData` and check its contents:

```{r load, warning = FALSE, message = FALSE, results = 'hide', eval = FALSE}
# Loading the data from SeuratData
InstallData("pbmc3k")
data(pbmc3k)

# We need to update the dataset to the latest Seurat version
pbmc3k <- UpdateSeuratObject(pbmc3k)

# Checking the contents of the object
pbmc3k
colnames(pbmc3k@meta.data)
```

```{r pbmc3k_res, echo = FALSE}
download.file("https://cf.10xgenomics.com/samples/cell/pbmc3k/pbmc3k_filtered_gene_bc_matrices.tar.gz", destfile = "pbmc3k_filtered_gene_bc_matrices.tar.gz")
untar("pbmc3k_filtered_gene_bc_matrices.tar.gz")
pbmc3k = Read10X(data.dir = "filtered_gene_bc_matrices/hg19/")
pbmc3k = CreateSeuratObject(counts = pbmc3k, project = "pbmc3k", min.cells = 3, min.features = 200)
unlink(c("pbmc3k_filtered_gene_bc_matrices.tar.gz","filtered_gene_bc_matrices"), recursive = TRUE)

pbmc3k
```

```{r pbmc3k_metadata, echo = FALSE}
colnames(pbmc3k@meta.data)
```

Since the object seems to contain a 'data' layer as well as cell annotations, and we are unsure to what extent is has been preprocessed and processed, we will start by testing our first function from `RightSeuratTools`, [Right_DietSeurat](https://alexis-varin.github.io/RightSeuratTools/reference/Right_DietSeurat.html), which is a reworked version of [DietSeurat](https://satijalab.org/seurat/reference/dietseurat), and will remove all the layers and slots from the object, leaving only the counts layer from the RNA assay. We will also remove all meta.data columns except for the 'orig.ident' metadata to trully start anew. [Right_DietSeurat](https://alexis-varin.github.io/RightSeuratTools/reference/Right_DietSeurat.html) is highly customizable, check the documentation for more information:

```{r right_dietseurat, warning = FALSE, message = FALSE, results = 'hide'}
# Reducing the object to the counts layer and the 'orig.ident' metadata
pbmc <- Right_DietSeurat(pbmc3k, idents = "orig.ident")

# Checking the contents of the object
pbmc
colnames(pbmc@meta.data)
```

```{r pbmc_res, echo = FALSE}
pbmc
```

```{r pbmc_metadata, echo = FALSE}
colnames(pbmc@meta.data)
```

## Data preprocessing

### Quality control and cell filtering

We will now move on to the preprocessing steps. We start by calculating the percentage of mitochondrial genes and plot these results alongside the number of genes (nFeature_RNA) and counts (nCount_RNA) per cell:

```{r mito, warning = FALSE, message = FALSE, fig.align = 'center', fig.width = 8, fig.height = 7}
# Calculate percentage of mitochondrial genes
pbmc[["percent.mt"]] <- PercentageFeatureSet(pbmc, pattern = "^MT-")

# Plot QC
ggplot(pbmc@meta.data, aes(x = nCount_RNA, y = nFeature_RNA, col = percent.mt)) + 
  geom_point(size=0.5) + 
  scale_color_gradientn(colors=c("black","blue","green3","yellow3","red3")) +
  ggtitle("pbmc QC metrics") +
  labs(x = "Log10(counts)", y = "Log10(number of genes)", col = "% mito") +
  scale_y_log10() +
  scale_x_log10() +
  theme_bw() +
  theme(panel.border = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line = element_line(colour = "black"))
```

Based on the plot above, we will filter out cells with more than 10% mitochondrial genes as well as cells with less than 400 or more than 2500 genes:

```{r filter, warning = FALSE, message = FALSE, results = 'hide'}
# Filter cells
pbmc <- subset(pbmc, subset = nFeature_RNA > 400 &
                              nFeature_RNA < 2500 &
                              percent.mt < 10)

# Checking the contents of the object
pbmc
```

```{r pbmc_filter_res, echo = FALSE}
pbmc
```

### Normalization

Now that we have filtered the cells, we will move on to the next step, which is normalizing and scaling the data and identifying highly variable genes. For convenience, we use the same parameters as in the `Seurat` tutorial:

```{r norm_hvg_scale, warning = FALSE, message = FALSE}
# Normalizing the data with default parameters
pbmc <- NormalizeData(pbmc)

# Find highly variable genes with default parameters
pbmc <- FindVariableFeatures(pbmc)

# Scaling all genes, by default it only scales the variable features
pbmc <- ScaleData(pbmc, features = rownames(pbmc))
```

### Dimensionality reduction

Next, we will perform dimensionality reduction and unsupervised clustering. We again use the same parameters as in the `Seurat` tutorial:

```{r dimred, warning = FALSE, message = FALSE, results = 'hide'}
# Perform PCA
pbmc <- RunPCA(pbmc)

# Clustering
pbmc <- FindNeighbors(pbmc, dims = 1:10)
pbmc <- FindClusters(pbmc, resolution = 0.5)

# UMAP
pbmc <- RunUMAP(pbmc, dims = 1:10)
```

Finally, we will plot the UMAP, which concludes the preprocessing steps:

```{r umap, warning = FALSE, message = FALSE, fig.align = 'center', fig.width = 8, fig.height = 7}
# Plot UMAP
DimPlot(pbmc, reduction = "umap", pt.size = 0.5)
```

Interestingly, compared to the `Seurat` tutorial, we are missing the platelet cluster, which is likely due to the fact that we have a more drastic cutoff for the number of genes per cell, at 400 versus 200. This is a good example of how the preprocessing steps can influence the clustering results.

## Downstream analysis

### Markers

A growing number of methods exist to label single-cell clusters, from using reference datasets with `SingleR` package or [Azimuth](https://azimuth.hubmapconsortium.org/), to querying ChatGPT with `GPTCelltype` package, via using canonical markers from the scientific literature. The vast majority of these methods need a set of genes as input, and determining the most differentially expressed genes (DEG) in each cluster is a common first step.

For this purpose, we are going to use our second function from `RightSeuratTools`, [Find_Annotation_Markers](https://alexis-varin.github.io/RightSeuratTools/reference/Find_Annotation_Markers.html), which is a wrapper function around [FindMarkers](https://satijalab.org/seurat/reference/findmarkers). It is designed by default to conveniently output a character vector of the top 5 markers, chosen based on the highest average log fold change in genes expressed in at least 25% of cells in each cluster, and with mitochondrial, ribosomal and non-coding genes excluded, in order to maximize the chances of finding canonical markers in an unsupervised manner. It also provides many more parameters which can be used to tailor the function to your specific needs:

```{r DEG, warning = FALSE, message = FALSE, results = 'hide'}
# Find markers
annotation.markers <- Find_Annotation_Markers(pbmc)

# Display markers for each cluster, only for visualization in the vignette
j = 1
for (i in 1:length(levels(Idents(pbmc)))) {
  cat("Cluster", i-1, ":", annotation.markers[j:(j+4)],"\n")
  j = j + 5
}
```

```{r DEG_res, echo = FALSE}
annotation.markers <- Find_Annotation_Markers(pbmc)
```

```{r DEG_display, echo = FALSE}
j = 1
for (i in 1:length(levels(Idents(pbmc)))) {
  cat("Cluster", i-1, ":", annotation.markers[j:(j+4)],"\n")
  j = j + 5
}
```

These markers may then be directly used for plotting. Since the `Seurat` tutorial uses [DoHeatmap](https://satijalab.org/seurat/reference/doheatmap), we will compare it to its equivalent in `RightSeuratTools`, [Cell_Heatmap](https://alexis-varin.github.io/RightSeuratTools/reference/Cell_Heatmap.html), which is reworked using `ComplexHeatmap` instead of `ggplot2`:

```{r heatmap, warning = FALSE, message = FALSE, fig.align = 'center', fig.width = 8, fig.height = 8}
# We will disable some parameters to be as close as possible from Seurat's DoHeatmap
Cell_Heatmap(pbmc,
             features = annotation.markers,
             cluster.features = FALSE,
             show.idents.legend = FALSE)
```

Let's compare it to `Seurat`'s default heatmap function:

```{r doheatmap, warning = FALSE, message = FALSE, fig.align = 'center', fig.width = 8, fig.height = 8}
DoHeatmap(pbmc,
          features = annotation.markers) + 
  NoLegend()
```

While both heatmaps look similar aside from colors, which is on purpose with default parameters, [Cell_Heatmap](https://alexis-varin.github.io/RightSeuratTools/reference/Cell_Heatmap.html) trully shines in complexity by offering more customization options, such as the possibility to cluster features, apply k-means, or split identities by another metadata...

### Cell annotation

We will now annotate each cluster with its corresponding cell type and display the UMAP:

```{r annotation, warning = FALSE, message = FALSE, fig.align = 'center', fig.width = 8, fig.height = 8}
# Annotate clusters
new.cluster.ids <- c("Naive CD4 T", "CD14+ Mono", "Memory CD4 T",
                     "B", "CD8 T", "FCGR3A+ Mono", "NK", "DC")

names(new.cluster.ids) <- levels(pbmc)
pbmc <- RenameIdents(pbmc, new.cluster.ids)

# Add metadata to Seurat object
pbmc@meta.data$named_clusters <- pbmc@active.ident

# Plot UMAP with cell types
DimPlot(pbmc, reduction = "umap", label = TRUE, pt.size = 0.5) +
  NoLegend()
```

### Cell proportion

Following cell annotation, one often want to visualize the proportion of each cell type. While using [table](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/table) on the identities gives this information, it is often tedious to organize these data for plotting; `RightSeuratTools` introduces [Barplot_Cell_Proportion](https://alexis-varin.github.io/RightSeuratTools/reference/Barplot_Cell_Proportion.html), which automates cell proportion and conveniently displays a bar plot of these cell fractions, with the possibility to group and/or split based on other metadata, and many other parameters:

```{r proportion, warning = FALSE, message = FALSE, fig.align = 'center', fig.width = 6, fig.height = 6}
Barplot_Cell_Proportion(pbmc,
                        percent = FALSE)
```

### Other visualizations

We will conclude by showing the last two visualization functions; [DotPlot_Heatmap](https://alexis-varin.github.io/RightSeuratTools/reference/DotPlot_Heatmap.html), which is a reworked version of [DotPlot](https://satijalab.org/seurat/reference/dotplot), also built from `ComplexHeatmap` instead of `ggplot2`, and [Grid_VlnPlot](https://alexis-varin.github.io/RightSeuratTools/reference/Grid_VlnPlot.html), which is a stacked version of [VlnPlot](https://satijalab.org/seurat/reference/vlnplot) in a square grid, both also offering many more options than their `Seurat` counterparts:

```{r dotplot, warning = FALSE, message = FALSE, fig.align = 'center', fig.width = 8, fig.height = 10}
# We will disable some parameters to be as close as possible from Seurat's DotPlot
# Due to the number of features, we will also lower dots size and flip the axis
DotPlot_Heatmap(pbmc,
                features = annotation.markers,
                dots.size = 2,
                cluster.features = FALSE,
                cluster.idents = FALSE,
                rotate.axis = TRUE)
```

Comparing it to `Seurat`'s default dotplot function:

```{r dotplot_seurat, warning = FALSE, message = FALSE, results = 'hide', fig.keep = 'none'}
DotPlot(pbmc,
        features = annotation.markers) + 
  RotatedAxis() + 
  coord_flip()
```

```{r dotplot_seurat_display, echo = FALSE, fig.align = 'center', fig.width = 8, fig.height = 10}
DotPlot(pbmc,
        features = annotation.markers) + 
  RotatedAxis() + 
  coord_flip() + 
  theme(plot.background = element_rect(fill = "white"))
```

And finally, the grid violin plot:

```{r grid_vln, warning = FALSE, message = FALSE, fig.align = 'center', fig.width = 8, fig.height = 8}
Grid_VlnPlot(pbmc,
             features = annotation.markers)
```


## Going further

### Advanced data visualization

While we have shown the default parameters of most of the functions of `RightSeuratTools`, there are many more options available in each function; to explore the use cases of most of these, and to see how far they can be customized, check the [advanced data visualization vignette](https://alexis-varin.github.io/RightSeuratTools/articles/Visualization.html).

### Gene signatures from GSEA pathways

Another interesting feature of `RightSeuratTools` is the possibility to extract genes from pathways in the Gene Set Enrichment Analysis (GSEA) database to create and visualize signatures. Check the [gene signatures from GSEA pathways vignette](https://alexis-varin.github.io/RightSeuratTools/articles/GSEA.html) to learn more about the various usages of this function.
