---
title: "Tutorial"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

This tutorial will guide you through the main functions of the `RightSeuratTools` package. The package is designed to provide complementary tools to the `Seurat` package, such as plotting, data manipulation, and more. The package is still under development, and new functions will be added in the future.

# Installation

You can install the `RightSeuratTools` package from GitHub using the `devtools` package. If you don't have the `devtools` package installed, you can install it using the following command:

> install.packages("devtools")

Once you have `devtools` installed, you can install the `RightSeuratTools` package using the following command:

> devtools::install_github("Alexis-Varin/RightSeuratTools")

# Loading the package

```{r}
library(RightSeuratTools)
```

# Data loading and preprocessing

For this tutorial, we will use the `pbmc3k` dataset from the `SeuratData` package. The `pbmc3k` dataset contains single-cell RNA-seq data from peripheral blood mononuclear cells (PBMCs). We will load the dataset and preprocess it using the `Seurat` package.

```{r}
library(Seurat)
library(SeuratData)

# Load the pbmc3k dataset
InstallData("pbmc3k")
data(pbmc3k)
pbmc3k = UpdateSeuratObject(pbmc3k)
pbmc3k

# To be continued...
```
